---
title: SPARKLE Training Course
subtitle: Fitting Models - Bayesian Inference
format:
  training-html: default
presenter: "Dr Eamon Conway"
contributors: [Dr Eamon Conway]
---

## Bayesian inference
Bayesian inference is an alternative to inferring model fits from data.
With maximum likelihood we have worked off the assumption that there is a true known value for the parameters and that the data varies.
That is, we are interested with $p(D|\theta$).
In direct contrast, for Bayesian inference we have that the data is known and we are interested in the probability distribution of the parameters, $p(\theta|D)$.
This is a subtle yet important point that has led to constant confusion and arguments.
The two points of view are also related through Bayes' Theorem.

::: {.callout title="Bayes' Theorem"}
In Bayesian inference, we use Bayes Theorem, 
$$
p(\theta|D) = \frac{p(D|\theta) p(\theta)}{p(D)},
$$
to convert our knowledge of how the data, $D$, was generated given the known parameters, $\theta$, into what our data says about the parameters, $p(\theta|D)$.
:::

### A simple example
Lets do linear regression in a Bayesian sense. 
Assume that we have measured the following data.

```{r echo=FALSE}
data <- (function() {
    N_samples <- 20
    X1 <- rnorm(N_samples, mean = 3, sd = 3)
    X2 <- rnorm(N_samples, mean = 10, sd = 4)
    return(tibble::tibble(
        X1 = X1,
        X2 = X2,
        Y = rnorm(N_samples, mean = 2 * X1 + 5, sd = 1)
    ))
})()

data |>
    tidyr::pivot_longer(cols = c("X1", "X2"), names_to = "Variable", values_to = "Measured") |>
    ggplot2::ggplot(ggplot2::aes(x = Measured, y = Y)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Variable, scale = "free")
```

This is 20 observed sampes, and I am going to assume that it is from a model such that, 
$$
D_i \sim \mathcal{N}(\mu = ax_1 + bx_2 + c, 1),
$$
where $x_i$ are measured covariates and $a$, $b$, and $c$ are parameters to be inferred.
Therefore we have a likelihood of,
$$
p(D|a,b,c) = \prod_{i=1}^{20} \mathcal{N}(D_i|ax_1 + bx_2 + c,1).
$$
If we were to look at a maximum likelihood approach we would now just determine the best parameters of $a$, $b$, and $c$. 
However, as we are doing Bayesian inference we want,
$$
p(a,b,c|D) \propto p(D|a,b,c)p(a,b,c),
$$
so we must first define what our prior for $p(a,b,c)$ is. 

For our purposes I am going to assume an uninformative prior,
$$
p(a,b,c) \propto 1
$$
You might notice this is not a distribution, which is okay, as long as the posterior is a distribution. 
When a prior is not a distribution we call it improper.

We are now in a position to infer our parameters, as we can state that our posterior is,
$$
p(a,b,c|D) \propto \prod_{i=1}^{20} \mathcal{N}(D_i|ax_1 + bx_2 + c,1).
$$
If the mathematics is not too complicated we may be able to determine the constant of proportionality, but this is often hard.
However, we have just learnt that there is a sampling technique that allows us to create a Markov chain with stationary distribution equal to our distribution of interest.

This is a Markov Chain Monte Carlo that constructs a sampling distribution that is the posterior distribution.
```{r echo=TRUE, filename="linear_regression.r"}
initial_samples <- list(
    c(a = 1, b = 1, c = 0),
    c(a = 10, b = 10, c = 10),
    c(a = 5, b = 4, c = 1),
    c(a = -1, b = -2, c = -3)
)
number_samples <- 4000
proposal_sd <- 2^-3

log_posterior <- function(x) {
    a <- x[1]
    b <- x[2]
    c <- x[3]
    mu <- a * data$X1 + b * data$X2 + c
    return(sum(dnorm(x = data$Y, mean = mu, sd = 1, log = TRUE)))
}

transition_kernel <- function(current) {
    proposed <- current + rnorm(3, 0, proposal_sd)
    log_acceptance_criteria <- log_posterior(proposed) +
        sum(dnorm(current, proposed, proposal_sd, log = TRUE)) -
        log_posterior(current) -
        sum(dnorm(proposed, current, proposal_sd, log = TRUE))
    if (log(runif(1, 0, 1)) > log_acceptance_criteria) {
        proposed <- current
    }
    return(proposed)
}

generate_mcmc_chain <- function(initial_sample) {
    return(
        purrr::accumulate(
            .x = 1:number_samples,
            .f = ~ transition_kernel(.x),
            .init = initial_sample
        )
    )
}

mcmc_chains <- purrr::map(
    .x = initial_samples,
    .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
        dplyr::mutate(Iteration = dplyr::row_number())
) |>
    dplyr::bind_rows(.id = "Chain") |>
    tidyr::unnest_wider(Sample) |>
    tidyr::pivot_longer(cols = c("a", "b", "c"), values_to = "Sample", names_to = "Variable")


mcmc_chains |>
    dplyr::filter(Iteration > 1000) |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Sample, color = Chain)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Variable)
```
We can use the samples to from the MCMC algorithm to estimate the shape of our posterior. 
```{r echo=TRUE, filename="plot_samples.R"}
mcmc_chains |>
    dplyr::filter(Iteration > 1000) |>
  ggplot2::ggplot(ggplot2::aes(x = Sample, color = Chain)) + 
  ggplot2::geom_histogram(ggplot2::density(),bins = 30)
```
An alternative to plotting samples from the posterior is to obtain a point estimate (these estimators minimise a loss function, see [Bayes' Estimators](https://en.wikipedia.org/wiki/Bayes_estimator)). 
When we were investigating maximum likelihood, we obtained a point estimate that was the maximum probability.
We can do the same and determine the maximum of the posterior. 
An alternative that I will demonstrate here is the mean of the posterior distribution. 

```{r echo=TRUE, filename="sliding_mean_chains.r"}
mcmc_chains |>
    dplyr::group_by(Variable, Chain) |>
    dplyr::arrange(Iteration, .by_group = TRUE) |>
    dplyr::mutate(Mean = slider::slide_dbl(Sample, mean, .before = 499)) |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Mean, color = Chain)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Variable)
```
The above figure demonstrates that as we obtain more samples from the posterior distribution that the expected value of our distribution stabilises.

## Exercises
1) 


## Futher reading
This topic barely touches on the approaches used in Bayesian inference. 
What is important to remember is that all of Bayesian inference is based around the posterior.
All of the methods you will find in the literature attempt to sample more efficiently from the posterior (for example Hamiltonian Monte Carlo), accurately approximate the posterior (for example INLA), or are methods to determine the best point estimate from the posterior. 
There are also approaches, such as using [conjugate priors](https://en.wikipedia.org/wiki/Conjugate_prior), which result in known posterior distributions.


<!-- 

An alternative, yet equally valid parameterisation, is to use log-odds ratio instead of probability in the binomial distribution.
By taking our posterior and defining the transformation $\phi = \log \theta - \log(1-\theta)$, our posterior can be expressed as, -->

