---
title: SPARKLE Training Course
subtitle: Fitting Models - Markov Chain Monte Carlo
format:
  training-html: default
presenter: "Dr Eamon Conway"
contributors: [Dr Eamon Conway"]
---
## Introduction
Samping from probability distributions is sometimes a complicated task. 
For common distributions there are known algorithms. 
For example the Box-Muller transform is known to generate random samples from a normal distribution, by sampling uniform random numbers. 
However, for more general distributions we must construct a sampling algorithm. 
A general purpose algorithm that is used widely is Markov Chain Monte Carlo (MCMC), along with its variants. 
In this module we review the neccesary elements of MCMC so that we can later use this algorithm for [Bayesian Inference](bayesianinference.qmd). 

## Overview of Monte Carlo methods

Monte Carlo methods are a broad range of techniques associated with approximating solutions to mathematical problems through repeated sampling (hence the name after the famous casino).
It is often likened to throwing darts at a problem and seeing which ones manage to stick.

To exemplify the a Monte Carlo approach, lets consider approximating the area of a circle with unit diameter.

  - As a first step, lets pretend that this is actually a really tough problem and we do not know the answer is $\frac{\pi}{4}$. 
  - For the second step, if the circle has unit diameter then we can place the circle entirely within a square with unit side lengths.
  - Next, lets throw 10000 darts at our square!

```{r }
dart_throw_locations <- tibble::tibble(
    x = runif(10000, -0.5, 0.5),
    y = runif(10000, -0.5, 0.5)
)
dart_throw_hits <- dart_throw_locations |>
    dplyr::mutate(Hit = (x^2 + y^2) < (0.5)^2)

ggplot2::ggplot(dart_throw_hits, ggplot2::aes(x = x, y = y, color = Hit)) +
    ggplot2::geom_point()

area <- dart_throw_hits |>
    dplyr::summarise(n = dplyr::n(), .by = tidyselect::all_of("Hit")) |>
    dplyr::mutate(proportion = n / sum(n))
print(area)
print(pi / 4)
```
The above code emulates the $x$ and $y$ location of each dart throw on our unit square. 
By visualising these plots over our circle, we can observe that some of the darts have succeeded in being inside the circle and some have not. 
We know that those darts had equal probability of being anywhere inside that square.
Therefore, if we were to count the number of darts that were inside of the circle and compare that to the number that were outside of the circle, we are approximating the percentage area of the square that the circle takes up. 
Due to the fact that the square has an area of 1 unit$^2$ we are approximating exactly the area of the circle (and not some scalar multiple of the circle).

What does this mean?
We can use random sampling to help answer our questions.
We can use an Accept-Reject step to take something that is similar to what we want and turn it into exactly what we want. 


## Overview of Markov Chains
::: {.callout title="Discrete Markov Chains"}
A discrete Markov Chain is a sequence of random variables that satisfy the Markov property.
That is, the next random variable in the sequence only depends upon the current state,
$$
X_{n+1} \sim P(\cdot | X_n)
$$
where $X_n$ is the $n$-th random variable in the sequence and $P(\cdot | X_n)$ is the proposal distribution.
:::

::: {.callout title="Detailed Balance"}
$$
\pi(x)p(x'|x) = \pi(x')p(x|x')
$$
where $\pi(x)$ is the stationary distribution of the Markov Chain and $p(x|x')$ is the transition probability from $x'$ to $x$.
The flow out of $x$ into $x'$ is balanced by the flow from $x'$ into $x$.
:::

## Markov Chain Monte Carlo 

Can you Monte Carlo your way to a Markov Chain with known stationary distribution?

Markov Chain Monte Carlo allows us to sample from the probability distribution $\pi(x)$, which may not have been possible before.

::: {.callout-warning}
If you can sample exactly from the distribution, you do not need to use Markov Chain Monte Carlo.
For example, if you want to sample from a normal distribution, just sample from the normal distribution.
:::

It would be extremely helpful if we could easily construct the transition probabilities of a Markov Chain with a known stationary distribution.
If this was the case then we could generate a large number of samples from our Markov Chain and the stationary distribution of these samples will be samples from our distribution of interest.
This is the main premise behind Markov Chain Monte Carlo, where we use a Monte Carlo approach to alter a simpler Markov Chain into one with a known stationary distribution.

To construct the transition probabilities we decompose $p(x'|x)$ into an accept-reject step.
Namely,
$$
p(x'|x) = q(x'|x)A(x', x),
$$
where $q(x|x')$ is the probability density of the proposal distribution and $A(x', x)$ is the acceptance criteria (note that it is both a function of the current state and the updated state).
The accept-reject step consists of proposing a new sample,
$$
x' \sim Q(\cdot|x)
$$
where $Q(\cdot|x)$ is the proposal distribution, followed by checking if we accept the new sample.
We can now ask what acceptance criteria will ensure that the Markov Chain satisfies detailed balance with a stationary distribution $\pi(x)$.

::: {.callout title="Metropolis-Hastings Acceptance Criteria"}
$$
A(x', x) = \min \left(1, \frac{\pi(x')q(x| x')}{\pi(x)q(x' | x)}\right)
$$
where $q(x'|x)$ is probability density of the proposal distribution of $x'$ given a known $x$.
:::

The Metropolis-Hastings Algorithm is a way to construct a Markov Chain with known stationary distribution.
This algorithm breaks up the sampling procedure into two steps, the proposal step and the acceptance step.
By using an acceptance step that satisfies detailed balance, we can ensure that the stationary distribution of the algorithm is the probability distribution we want to sample from.
This is beneficial when you do not have another way to sample from the required distribution.

The full algorithm for Markov Chain Monte Carlo is:

1) Start with an initial sample $x$.
2) Generate a proposed sample $x' \sim Q(\cdot | x)$
3) Accept sample with probability $A(x', x)$.
4) Repeat steps 2-4 as desired.

This will generate a sequence of samples that have the stationary distribution $\pi(x)$.

### Example
Here we provide a simple implementation of Markov Chain Monte Carlo.
Consider we want to sample from a normal distribution with mean $40$ and standard deviation of $2$.
We can obviously sample directly from that distribution using known techniques, but lets construct a Markov Chain such that
$$
\pi(x) = \mathcal{N}(x|\mu = 40, \sigma = 2).
$$

```{r echo=T, filename="simple_mcmc.r"}
number_samples <- 10000
initial_samples <- c(1.0, 5.0, -10.0, 40.0)
mu <- 40
sd <- 2

transition_kernel <- function(current) {
    proposed <- current + rnorm(1, 0, 1)
    acceptance_criteria <- dnorm(proposed, mu, sd) * dnorm(current, proposed, 1) / (dnorm(current, mu, sd) * dnorm(proposed, current, 1))
    if (runif(1, 0, 1) > acceptance_criteria) {
        proposed <- current
    }
    return(proposed)
}

generate_mcmc_chain <- function(initial_sample) {
    return(
        purrr::accumulate(
            .x = 1:number_samples,
            .f = ~ transition_kernel(.x),
            .init = initial_sample
        )
    )
}

mcmc_chains <- purrr::map(
    .x = initial_samples,
    .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
        dplyr::mutate(Iteration = dplyr::row_number())
) |>
    dplyr::bind_rows(.id = "Chain")

mcmc_chains |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Chain)

mcmc_chains |>
    dplyr::filter(Iteration > 250) |>
    ggplot2::ggplot(ggplot2::aes(x = Sample)) +
    ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
        bins = 50
    ) +
    ggplot2::geom_function(fun = dnorm, args = list(mu, sd))
```

There are hyperparmeters within the above algorithm that should be understood, the most important of which are associated with the proposal distribution.
By calibrating the proposal distribution it is possible to end up with fast (or slower) convergence to the stationary distribution.

```{r echo=F}
number_samples <- 200
initial_sample <- 1
mu <- 40
sd <- 2
proposal_sds <- 2^(seq(0, 3))

transition_kernel <- function(current, proposal_sd) {
    proposed <- current + rnorm(1, 0, proposal_sd)
    acceptance_criteria <- dnorm(proposed, mu, sd) * dnorm(current, proposed, proposal_sd) / (dnorm(current, mu, sd) * dnorm(proposed, current, proposal_sd))
    if (runif(1, 0, 1) > acceptance_criteria) {
        proposed <- current
    }
    return(proposed)
}

generate_mcmc_chain <- function(proposal_sd) {
    return(
        purrr::accumulate(
            .x = 1:number_samples,
            .f = ~ transition_kernel(.x, proposal_sd),
            .init = initial_sample
        )
    )
}

mcmc_chains <- purrr::map(
    .x = proposal_sds,
    .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
        dplyr::mutate(Iteration = dplyr::row_number(), sd = .x)
) |>
    dplyr::bind_rows()

mcmc_chains |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~sd)
```

In our simple example the only hyperparameter within the proposal distribution is the proposal variance.
Figure XX demonstrates the effect of changing the proposal variance.
The time to reach the stationary distribution is obviously affected by the chosen variance and there will be an optimal choice (that is problem dependent).
It is important to note that as the proposal variance increases, we also observe an increased in the auto-correlation between samples.
This is partly attributed to an increased probability of the proposed sample being rejected.
Having a wider proposal distribution has made it more likely that samples are rejected.

::: {.callout-warning}
The implementation presented above is prone to numerical issues.
Numerical methods in statistics often have to deal with underflow and overflow, as such it is better to do computations in log-space.
:::

::: {.callout title="Proportionality"}
A very helpful property of the acceptance criteria is that we only need to know the stationary distribution up to a constant of proportionality.
Therefore, if we have a function,
$$
f(x) \propto \pi(x),
$$
we can still construct the Markov Chain with stationary distribution $\pi(x)$ by
$$
A(x', x) = \min \left(1, \frac{f(x')q(x| x')}{f(x)q(x' | x)}\right).
$$
:::

The majority of methods that speed up the convergence of Markov Chain Monte Carlo focus on choosing better proposal distributions.
A small and incomplete list of these methods include:

- Langevin Monte Carlo
- Hamiltonian Monte Carlo

## Numerical stability
One of the most integral parts of computational math is numerical stability.
We have to ensure that our algorithms are stable, at least in the regime we intend to use them.
For the Markov Chain Monte Carlo algorithm numerical stability can be improved by working with log probabilities.
If you do not work with log probabilities your implementation will be subject to both numerical over- and under-flow, depending on the normalisation factor that you ignore.

To develop the algorithm with log probabilities we must determine the acceptance criteria on the log scale.
Taking the Metropolis-Hastings acceptance criteria we state,
$$
\log A(x', x) = \log \min \left(1, \frac{\pi(x')q(x| x')}{\pi(x)q(x' | x)}\right) = \min\left(0, \log \left(\frac{\pi(x')q(x| x')}{\pi(x)q(x' | x)}\right)\right),
$$
due to the monotonically increasing nature of the $\log$ function.
Therefore we can work entirely with log probabilities as,
$$
\log A(x', x) = \min\left(0, \log \pi(x') + \log q(x| x') - \log \pi(x) - \log q(x' | x)\right).
$$
The accept-reject step of the algorithm, is now altered to check if,
$$
\log U < \log \pi(x') + \log q(x| x') - \log \pi(x) - \log q(x' | x),
$$
where $U$ is uniformly distributed between 0 and 1.

```{r echo=T, filename="stable_mcmc.r"}
number_samples <- 10000
initial_samples <- c(1.0, 5.0, -10.0, 40.0)
mu <- 40
sd <- 2

transition_kernel <- function(current) {
    proposed <- current + rnorm(1, 0, 1)
    log_acceptance_criteria <- dnorm(proposed, mu, sd, log = TRUE) +
        dnorm(current, proposed, 1, log = TRUE) -
        dnorm(current, mu, sd, log = TRUE) -
        dnorm(proposed, current, 1, log = TRUE)
    if (log(runif(1, 0, 1)) > log_acceptance_criteria) {
        proposed <- current
    }
    return(proposed)
}

generate_mcmc_chain <- function(initial_sample) {
    return(
        purrr::accumulate(
            .x = 1:number_samples,
            .f = ~ transition_kernel(.x),
            .init = initial_sample
        )
    )
}

mcmc_chains <- purrr::map(
    .x = initial_samples,
    .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
        dplyr::mutate(Iteration = dplyr::row_number())
) |>
    dplyr::bind_rows(.id = "Chain")

mcmc_chains |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Chain)

mcmc_chains |>
    dplyr::filter(Iteration > 250) |>
    dplyr::slice(which(Iteration %% 20 == 0)) |>
    ggplot2::ggplot(ggplot2::aes(x = Sample)) +
    ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
        bins = 50
    ) +
    ggplot2::geom_function(fun = dnorm, args = list(mu, sd))
```

### Limitations

There are a number of limitations to the application of MCMC. 

- The distribution function, $\pi(x)$, must be evaluated many times. 
Depending upon the implementation, the parameter space, and your data, this function may be slow to evaluate and thus very limiting.
If you cannot evaluate your distribution function in a reasonable amount of time, your Markov chain may not reach steady state and you will not be sampling from the distribution $\pi(x)$. 

- The samples from your Markov chain depend upon the previous samples, therefore, each sample will be dependent upon the samples that came before it. 
This similarity is [autocorrelation](https://en.wikipedia.org/wiki/Autocorrelation).
Autocorrelation poses problems for MCMC, notably, the parameter space may not be explored efficiently and the chain instead stays localised.
Samples are often thinned to overcome autocorrelation. 
Thinning involves removing all but the $n$th recurring sample. 
Therefore, if you want 1000 samples and thin so that you only retain every 5th sample you will have to run the chain for 5000 samples (this is before burn in is accounted for).

- The hyperparameters that are present in the Markov Chain Monte Carlo algorithm directly affect the convergence and stability of the algorithm.
Choosing hyperparameters incorrectly can make your implementation infeasible. 
There is a vast amount of literature on how to choose hyperparameters, but this is a guide only and is typically problem dependent. 

## Modern methods

Instead of writing your own MCMC sampler from scratch there are many freely available implementations online. 
These implementations have been thoroughly tested and used for many problems. 
To name just a few we have BUGS, JAGS, STAN and many many more implementations available within the R ecosystem. 
All of these samplers are creating markov chains so they follow the same general principles that were outlined in this section. 

## Exercises

1) Taking the normal distribution example in `stable_mcmc.r`, what happens if we only run the markov chain for 100 iterations?
    - Does each chain return the same value? 
    - Run `stable_mcmc.r` with `number_samples = 1000`. 
    What is the running average of each chains?
    When does the expected value stabilise?
    - Change the standard deviation of the proposal distribution, how does this affect the running average of each chain?
   
2) Implement markov chain monte carlo to sample from a Student's $t$-distribution where, 
$$
f(t) = \frac{\gamma(\frac{\nu+1}{2})}{\sqrt{\pi \nu \gamma(\frac{\nu}{2})}}\left(1 + \frac{t^2}{\nu}\right)^{-(\nu +1)/2},
$$
where $\nu$ = 3.
    - Show that the expected value of the distribution is zero
    - Show that the variance of the samples approximates the known variance, $\nu/(\nu - 2)$.


3) Using any markov chain monte carlo sampler you have made, use the `acf` function in R to plot the autocorrelation of samples. 
    - What do you notice about the autocorrelation?
    - Thin the samples, using the autocorrelation as a guide, and re-run `acf` on the thinned samples.
    What has happened to the autocorrelation?
    - What is the running expected value from the thinned samples?

4) Generate samples from 
$$
\pi(x) = \frac{1}{2}\mathcal{N}(x|10,1) + \frac{1}{2}\mathcal{N}(x|-10,1),
$$
where 
$$
\mathcal{N(x|\mu,\sigma^2) = \frac{1}{\sqrt{2 \pi\sigma^2}}\exp(-\frac{(x-\mu)^2}{2\sigma^2}),
$$
is the probability density function of the normal distribution. 
    - What do you notice about the samples? 
    Has each starting point finished at the final distribution? 
    - Are there any problems? 
    Why do you think they occurred?  


```{r echo=T, filename="student_t_sampler.r"}
number_samples <- 10000
initial_samples <- c(1.0, 5.0, -10.0, 40.0)
nu = 3

log_distribution = function(x) dt(x,nu,log=TRUE)

transition_kernel <- function(current) {
    proposed <- current + rnorm(1, 0, 1)
    log_acceptance_criteria <- log_distribution(proposed) +
        dnorm(current, proposed, 1, log = TRUE) -
        log_distribution(current) -
        dnorm(proposed, current, 1, log = TRUE)
    if (log(runif(1, 0, 1)) > log_acceptance_criteria) {
        proposed <- current
    }
    return(proposed)
}

generate_mcmc_chain <- function(initial_sample) {
    return(
        purrr::accumulate(
            .x = 1:number_samples,
            .f = ~ transition_kernel(.x),
            .init = initial_sample
        )
    )
}

mcmc_chains <- purrr::map(
    .x = initial_samples,
    .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
        dplyr::mutate(Iteration = dplyr::row_number())
) |>
    dplyr::bind_rows(.id = "Chain")

mcmc_chains |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Chain)

mcmc_chains |>
    dplyr::filter(Iteration > 250) |>
    # dplyr::slice(which(Iteration %% 20 == 0)) |>
    ggplot2::ggplot(ggplot2::aes(x = Sample)) +
    ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
        bins = 50
    ) +
    ggplot2::geom_function(fun = dt, args = list(nu))
```

```{r echo=T, filename="separated_modes.r"}
number_samples <- 10000
initial_samples <- c(1.0, 5.0, -10.0, 10.0)
mu_1 <- 10
mu_2 <- -10
sd <- 1

log_distribution = function(x){ log(dnorm(x, mu_1, sd, log = FALSE) + dnorm(x, mu_2, sd, log = FALSE))}

transition_kernel <- function(current) {
    proposed <- current + rnorm(1, 0, 1)
    log_acceptance_criteria <- log_distribution(proposed) +
        dnorm(current, proposed, 1, log = TRUE) -
        log_distribution(current) -
        dnorm(proposed, current, 1, log = TRUE)
    if (log(runif(1, 0, 1)) > log_acceptance_criteria) {
        proposed <- current
    }
    return(proposed)
}

generate_mcmc_chain <- function(initial_sample) {
    return(
        purrr::accumulate(
            .x = 1:number_samples,
            .f = ~ transition_kernel(.x),
            .init = initial_sample
        )
    )
}

mcmc_chains <- purrr::map(
    .x = initial_samples,
    .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
        dplyr::mutate(Iteration = dplyr::row_number())
) |>
    dplyr::bind_rows(.id = "Chain")

mcmc_chains |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration, y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Chain)

mcmc_chains |>
    dplyr::filter(Iteration > 250) |>
    ggplot2::ggplot(ggplot2::aes(x = Sample)) +
    ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)),
        bins = 50
    ) 
```